Incremental tactics

### 2012-05-08 doc/papers/LFMTP2012/paper.tex
paper

As programming languages develop more static analysis like rich type
systems, programs get safer as bugs are caught earlier.

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects. We have the ability to construct
objects out of other, and deconstruct them by case analysis:
\begin{align*}
  T &\gequal
  \lam x T \gor
  U \\
  U &\gequal
  F \gor
  \matchin U \Gamma \casenobar {\app {H} {\vec {\var x}}} U\ |\ \ldots\
  \case {\app {H} {\vec {\var x}}} U
\end{align*}

This language is un(i)typed, in that all \LF\ objects have the same
implicit type $\cst{obj}$: there is no static guarantee that terms
produced by a constant $\cst c : A_1\to A_2$ will have type $A_2$; it
is also non-terminating: we could declare a constant $\cst c : \cst
a\to \cst a = \lam x \app {\cst c} {\var x}$. We expose here only the
minimal building blocks, but other features could be added at will to
this language. You can think of defined constants as black boxes
taking and producing terms, and being able to call other black
boxes. We detail in section \ref{sec:implem} how to embed
\sysname{OCaml} as computation calculus.

Nevertheless, in order to maintain the well-typing of the generated
term, we want to maintain the invariant that arguments fed to these
defined constant as well as the computation result have the announced
type. We thus maintain \emph{dynamic typing} of computations.

\begin{figure*}
  \centering
  TODO
  \caption{Weak head-reduction}
  \label{fig:whr}
\end{figure*}

>>>>

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects. We have the ability to construct
objects out of other, and deconstruct them by case analysis:
\begin{align*}
  T &\gequal
  \lam x T \gor
  U \\
  U &\gequal
  F \gor
  \matchin U \Gamma C \\
  C &\gequal \casenobar P U \gor
  C\ \case P U \\
  P &\gequal
  \app H {\var x} \ldots\ {\var x}
\end{align*}

This language is un(i)typed, in that all \LF\ objects are viewed as
having the same implicit type $\cst{obj}$: there is no static
guarantee that terms produced by a constant $\cst c : A_1\to A_2$ will
have type $A_2$. It is also non-terminating: we could declare a
recursive constant $\cst c : \cst a\to \cst a = \lam x \app {\cst c}
{\var x}$. We expose here only the minimal building blocks of this
language, but really any language with algebraic datatypes would do
since we will be manipulating directly the first-order representation
of \LF\ objects. In fact, whatever the language is, you can think of
defined constants as mere black boxes taking and producing terms, and
being able to call other black boxes. We detail for example in section
\ref{sec:implem} how to embed \sysname{OCaml} as computation calculus.

Nevertheless, in order to maintain the well-typing of the generated
term, we want to maintain the invariant that each time a piece of term
goes in or out of a black box, it is checked against its specification
(\ie\ its type). We thus maintain \emph{dynamic typing} of
computations: arguments and results are type-checked at run-time.

\begin{align*}
  \gsubst {(\lam x T)} \sigma &= \lam x {\gsubst T\sigma} \qquad
  \text{if $\var x\notin\dom\sigma$}\\
  \gsubst {(\matchin U \Gamma C)} \sigma &=
  \matchin {\gsubst
    U\sigma} {\gsubst\Gamma\sigma} {\gsubst C\sigma} \\
  \gsubst {(\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
          x_n}}}} U)} \sigma &= {\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
          x_n}}}} \gsubst U \sigma} \quad \text{if $\var x_i\notin\dom\sigma$} \\
\end{align*}

\begin{figure*}
  \fbox{$\jcc\mr\Gamma T S A \mmr F$}
  \qquad
  {Computational cut}

  \begin{mathpar}
    \infer{
      \jm\mr\Gamma M A \mmr {M'}
      \and
      \jcc\mmr\Gamma{\gsubst T {\msubst x M}} S {A_2}\mmmr F
    }{
      \jcc\mr\Gamma{\lam x T}{\spinecons M S}{(\prd x {A_1} {A_2})}
      \mmmr F
    }

    \infer{
      \jcr\mr\Gamma U \mmr F
      \and
      \jm\mmr\Gamma F {\app {\cst a} S} \mmmr {F'}
    }{
      \jcc\mr\Gamma U \spinenil {\app {\cst a} S} \mmmr {F'}
    }
  \end{mathpar}

  \fbox{$\jcr\mr\Gamma U \mmr F$}
  \qquad
  {Computational reduction}

  \begin{mathpar}
    \infer{ }{
      \jcr\mr\Gamma{\app {\var x} S} \mr {\app {\var x} S}
    }

    \infer{
      \Sigma(c) = A; (\annsliceable\text{ or }\annnonsliceable)
    }{
      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
    }

    \infer{
      \Sigma(c) = A; \anndefined T
      \and
      \jcc\mr\Gamma T S A \mmr F
    }{
      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
    }

    \infer{
      \jcr\mr{\gconcat\Gamma{\Gamma'}} U \mmr {\app {H_i} {\spinecons
          {M_1}{\spinecons\ldots{M_{m_i}}}}}
      \and
      \jcr\mmr\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
          \msubst {x$_{m_i}$} {M_{m_i}}}} \mmmr F a
    }{
      \jcr\mr\Gamma{\matchin{U}{\Gamma'}{
          \casenobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
                \ldots\ {\var x_{1m_1}}}}}{U_1}
          |\ \ldots\
          \case{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
              \ldots\ {\var x_{nm_n}}}}}{U_n}
        }}\mmr F
    }

  \caption{Weak head-reduction}
  \label{fig:whr}
\end{figure*}
